---
title: "DataClean"
format: html
editor: visual
---

## 1. BCCWS

### 1.1 Data Clean

Cleaning the BCCWS is done in several step, which will be executed here using a single function. All that is needed manually is the start and end years of the analysis window.

```{r}

#specify start and end year
Y1 = 2008
Y2 = 2022

#Run the BCCWS cleaning scripts. 
#Your output will include the data you need for an analysis of trends. 
source("BCCWSClean.R")

```

In your environment you will have the following objects:

`in.BCCWS`: the clean data matrix

`events.BCCWS`: the events matrix used for zero-filling your data matrix

## 2. PSSS

### 2.1 Data Clean

A similar function as above is used to clean the PSSS data.

```{r}

#specify start and end year
Y1 = 2008
Y2 = 2022

#Run the BCCWS cleaning scripts. 
#Your output will include the data you need for an analysis of trends. 
source("PSSSClean.R")



```

There are 80 species that are monitored by both the BCCWS and PSSS:

```         
"COGO" "RNGR" "SBIG" "COME" "PALO" "BAEA" "BUFF" "SUSC" "DCCO" "WWSC" "GWGU" "MAMU" "AMWI" "HOGR" "HARD" "PECO" "MALL" "COLO" "BAGO" "LTDU" "RBME" "HERG" "GBHE" "RTLO" "WEGR" "PIGU" "GADW" "GWTE" "RBGU" "LESC" "BLSC" "NOPI" "HOME" "BRAN" "CACG" "PEFA" "NSHO" "MERL" "RNDU" "GRSC" "NOHA" "SNGO" "RHAU" "PBGR" "HEEG" "EUWI" "CANV" "OSPR" "BRAC" "CAGU" "COMU" "CAGO" "BOGU" "WEGU" "COHA" "SSHA" "EAGR" "ANMU" "GWFG" "AMCO" "GLGU" "CITE" "TUSW" "RUDU" "REDH" "CATE" "COTE" "KIEI" "YBLO" "RNPH" "CAAU" "POJA" "SOSH" "NOFU" "REPH" "CLGR" "BRPE" "SAGU" "AWPE" "IVGU"
```

NOTE: Not all of these will make the minimum data requirements.
