---
title: "03_DataAnalysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Data Analysis - Species Trends

## 3 International and National Assessments

A first assessment of coastal waterbird trends will be done for the full extent of the Salish Sea, and for each country separately. This is done on a species-by-species basis and is done without using a spatially explicit model to start. 

Run setup if starting a new environment.
```{r}

source("00_Setup.R")

#Manually Specify the start and end year of the analysis. These years should match those selected for the data cleaning scripts. 
Y1 = 2008
Y2 = 2022

```


### 3.1 International or National Assessment 

The first step is to run the analysis for the full extent of the Salish Sea, or for each country separately. Species are only included in an analysis if they meet the minimum data requirements, which are detailed below.  

We approached this analysis using a spatially varying coefficient approach (SVC, Gelfand et al. 2003) which allowed for information on abundance and trends to be shared between geographic locations. Specifically, we used a continuous space triangulated model mesh and a stochastic partial differential equation (SPDE) approach (Lindgren et al. 2022). Duration in hours was included as a covariate to account for differences in sampling effort. A first-order autoregressive model was used to account for temporal autocorrelation in counts based on the day of year (doy_idx). A random effect was included to account for Routes level effects.

The basic model formula is as follows:

Observation Count ~ DurationInHours + 
      f(kappa, model="iid", hyper=hyper.iid) + f(doy_idx, model = "ar1", hyper=prec.prior) 
    + f(alpha, model =spde)+ f(tau, model =spde)
    
Define your `site` as either "SalishSea", "BCCWS" or "PSSS".
       
```{r}

#Load your saved species data 
sp.data<-read.csv("Data/sp.data.csv")

#Create a list of species
sp.list<-unique(sp.data$SpeciesCode)
#manually select species
#sp.list <- c("AMWI", "GRSC") 

#Load your saved events data which is needed for zero-filling
events<-read.csv("Data/events.csv")

#Define your site. 
site <- "PSSS"

#Create a data frame to store the results. 
#This is the template used for the State of Canada's Birds and is required for upload into NatureCounts
source("OutputTables.R")

#Select the minimum data requirements for the analysis. The ones selected here were done follwoing the inspection of the International data set and likely work both the National anlysis. However, finer scale assessment may need to assess their suitaprovides bility. 

#Across routes with at least one detection, each species needs an overall abundance per year > 10, need to be detected in > one-half of the survey years, and on > 10 survey routes). Users may choose to adjust these parameters. 

min.abundance <- 50
min.years <- (Y2-Y1)/2
nroutes <- 10

#Set the distributional assumption. We will start with 'nbinomal' but may need to adjust this for some species if there is over dispersion. 

fam<-'nbinomial'
#fam<-'poisson'

#Now we can initiate the analysis.  
source("Analysis.R")

```

Note that the Dispersion Statistic file in the Output folder should be reviewed after the analysis. If the statistic is > 10 you should inspect the FitPlots in the Plot directory. In this case we will want to rerun using a different distributional assumption on the counts. This can be done by manually changing the `fam` to Poission and selecting these species to be rerun. 

**AMWI and GRSC manually removed due to poor model fit.

#Plot the results to a pdf for inspection. You will find the outputs in your Output/Plot folder with the name site_IndexPlot.pdf

```{r}

source("PlotTrends.R")

```
      


#check the sp.dat for PSSS. Plot the counts per year to check for outliers

```{r}

ggplot<-ggplot(sp.dat, aes(x=wyear, y=ObservationCount, color=SpeciesCode)) + geom_point() + facet_wrap(~SpeciesCode, scales="free_y") + theme_minimal() + theme(legend.position="none") + ggtitle("Counts per Year for PSSS") + theme(plot.title = element_text(hjust = 0.5))

```