---
title: "03_DataAnalysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Data Analysis - Species Trends

## 3 International and National Assessments

A first assessment of coastal waterbird trends will be done for the full extent of the Salish Sea, and for each country separately. This is done on a species-by-species basis and is done without using a spatially explicit model to start. 

Run setup if starting a new environment.
```{r}

source("00_Setup.R")

#Manually Specify the start and end year of the analysis
Y1 = 2008
Y2 = 2022

```


### 3.1 International or National Assessment 

The first step is to run the analysis for the full extent of the Salish Sea, or for each country separately. Species are only included in an analysis if they meet the minimum data requirements, which are detailed below.  

We approached this analysis using a spatially varying coefficient approach (SVC, Gelfand et al. 2003) which allowed for information on abundance and trends to be shared between geographic locations. Specifically, we used a continuous space triangulated model mesh and a stochastic partial differential equation (SPDE) approach (Lindgren et al. 2022). Duration in hours was included as a covariate to account for differences in sampling effort. A first-order autoregressive model was used to account for temporal autocorrelation in counts based on the day of year (doy_idx). A random effect was included to account for Routes level effects.

The basic model formula is as follows:

Observation Count ~ DurationInHours + 
      f(kappa, model="iid", hyper=hyper.iid) + f(doy_idx, model = "ar1", hyper=prec.prior) 
    + f(alpha, model =spde)+ f(tau, model =spde)
       
```{r}

#Load your saved species data 
sp.data<-read.csv("Data/sp.data.csv")
#Create a list of species
sp.list<-unique(sp.data$SpeciesCode) 
#remove NOFU from the sp.list
sp.list<-sp.list[sp.list!="NOFU"]

#Load your saved events data which is needed for zero-filling
events<-read.csv("Data/events.csv")

#Define your site. The option here are "SalishSea", "BCCWS" or "PSSS"
site <- "SalishSea"

#Create a data frame to store the results. 
#This is the template used for the State of Canada's Birds and is required for upload into NatureCounts
source("OutputTables.R")

#Select the minimum data requirements for the analysis. The once selected here are consistent with Ethier et al. (2020). Specifically, (i.e., species with > 85% zero counts across all routes with at least one detection, species that had a mean abundance per year < 5, and species that were detected in fewer than one-half of the survey years). Users may choose to adjust these parameters. 

min.abundance <- 5
min.years <- (Y2-Y1)/2
zero.count<-0.85

#Now we can initiate the analysis.  
source("Analysis.R")

```

