---
title: "03_DataAnalysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Data Analysis - SPDE

## 3 International and National Assessments

A first assessment of coastal waterbird trends will be done for the full extent of the Salish Sea, and for each country separately. This is done on a species-by-species basis and is done without using a spatially explicit model to start.

The first step is to run the analysis for the full extent of the Salish Sea, or for each country separately. Species are only included in an analysis if they meet the minimum data requirements, which are detailed below.

We approached this analysis using a spatially varying coefficient approach (SVC, Gelfand et al. 2003) which allowed for information on abundance and trends to be shared between geographic locations. Specifically, we used a continuous space triangulated model mesh and a stochastic partial differential equation (SPDE) approach (Lindgren et al. 2022). Duration in hours was included as a covariate to account for differences in sampling effort. A first-order autoregressive model was used to account for temporal autocorrelation in counts based on the day of year (doy_idx). A random effect was included to account for site level effects.

The basic model formula is as follows:

Observation Count \~ -1 + f(kappa, model="iid", hyper=hyper.iid) + f(doy_idx, model = "ar1", hyper=prec.prior) + f(alpha, model =spde)+ f(tau, model =spde)

\*Duration in hours is included as an offset in the INLA framework to model the expected count based on a difference in sampling effort.

Run setup if starting a new environment.

```{r}

source("00_Setup.R")

#Manually Specify the start and end year of the analysis. These years should match those selected for the data cleaning scripts. 
Y1 = 2008
Y2 = 2024

#Load your saved species data 
sp.data<-read.csv("Data/sp.data.csv")
#Load your saved events data which is needed for zero-filling
events<-read.csv("Data/events.csv")

```

### 3.1 Set Analysis Conditions

Here the user has several customization options.

-   Select the study area.

-   For a species-specific analysis, select the species you wish to analyse.

To view your options of species:
```{r sp}

species<-unique(sp.data$CommonName)
view(species)

```

-   For a guild-specific analysis, select the guild you wish to analyse (i.e., migration, diet, or family), and the species will be used as a random effect in the model.

To view your option of guilds:
```{r guild}

migration<-unique(sp.data$Migration)
view(migration)

diet<-unique(sp.data$Diet)
view(diet)

family<-unique(sp.data$family_name)
view(family)

```

-   Set the minimum data requirements.

-   Set the distributional family.

**Note: R is case sensitive so please type your variable names carefully and as specified below**

```{r}

#Define your `site` as either "SalishSea", "BCCWS" or "PSSS". 
area <- "SalishSea"

#Create a list of species using all available species in the shared dataset.
species.list<-"ALL"

#Or you can manually select species.
#species.list <- c("American Wigeon", "Common Loon", "Large Gull") 

#If you wish to do a guild-specific analysis, change to "Yes" and specify the `type` as either "migration", "diet", or "family". This will override the species list above. 

guild <- "No"
type <-"family"

#Select the minimum data requirements for the analysis. The ones selected here were done following the inspection of the International data set and likely work both the National analyses. However, finer scale assessment may need to assess their suitability. 

#Here, the minimum data required across sites with at least one detection, each species needs an overall abundance per year > 10, need to be detected in > one-half of the survey years, and on > 10 survey sites. Users may choose to adjust these parameters. 

min.abundance <- 10
min.years <- (Y2-Y1)/2
nsites <- 10

#Set the distributional assumption. Here we select 'nbinomal' but this may need to adjust if there is overdispersion. 

fam<-'nbinomial'
#fam<-'poisson'

```

### 3.2 Run the Analysis and Plot Output

```{r}

#Now we can initiate the analysis.  
source("Analysis.R")

#Plot the results to a pdf for inspection. 
#You will find the outputs in your Output/Plot folder with the name site_IndexPlot.pdf
source("PlotTrends.R")


```

Note that the Dispersion Statistic file in the Output folder should be reviewed after the analysis. If the statistic is \> 10 you should inspect the FitPlots in the Plot directory. In this case we will want to rerun using a different distributional assumption on the counts. This can be done by manually changing the `fam` to Poission and selecting these species to be rerun.

\*\*AMWI and GRSC manually removed from the International outputs due to poor model fit.

### 3.3 Combine and Plot Results

If you have run the analysis for all three regions (SalishSea, BCCWS, PSSS) you can now combine the results into a single plot and table for review.

```{r}

#You can only plot one trend type in the plots. 
#Select either End Point trend or Slope trend. 
type<-"End Point trend"

#Plot the results to a pdf for inspection. You will find the outputs in your Output/Plot folder with the name Full_TrendTable.csv
source("PlotInter&National.R")


```
